Unsupervised embeddings of organizations capture latent structure of scientific mobility

### 2020-03-08 paper/current/main.tex

Where the flux, $T_{ij}$, or number of authors co-affiliated with a pair organizations during the time period increases with the organization's sizes, $m_{i}$ and $m_{j}$, and decays as a function of their distance, $f(r_{ij})$.
We consider two distances, the geographic distance between the organizations as a baseline, and the \textit{embedding distance} defined as the cosine distance between the organization's vector in the trained embedding space.
We argue that the performance of a distance measure can be observed from their ability to explain expected mobility between organizations and to make predictions using the gravity model.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

The effective distance estimated by our embedding model explains and predicts scientific mobility best.
The amount of variance explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds for the variance of only domestic (Fig.~\ref{fig:supp:gravity_correlation}c) and international mobility flows (Fig.~\ref{fig:supp:gravity_correlation}d).
After fitting model parameters from the data, the embedding distance produces better predictions of expected flux than does geographic distance globally (Fig.~\ref{fig:gravity_performance}c--d) and for domestic (Fig.~\ref{fig:supp:gravity_correlation}e--f) and international mobility(Fig.~\ref{fig:supp:gravity_correlation}g--h).
Similarly, embedding distance outperforms ``diffusion''  distances such as those derived from calculating personalized-page rank scores (see supporting materials) on the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Geographic distance alone is a poor tool for understanding international scientific mobility---rather, by learning a distance directly from the data, it is possible to better explain, and better predict flux between organizations.

@@

%
% S1 TEXT
\paragraph*{S1 Text}
\label{si:text:ppr_dist}
{\bf Network-based personalized page rank distances.}

@@

%
% Figure - Geographic predictions, by model
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance_geo_bymodel.pdf}
	\caption{
		\textbf{For geographic distance, the power-decay gravity model is better.}
		Predictions of flux between organization pairs made using geographic distance the exponential gravity model (top), and the power-law decay model (bottom). 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
		Shown using all pairs of organization (\textbf{a-b}), domestic (\textbf{c-d}), and international only (\textbf{e-f}) mobility. 
		The power-decay model outperforms the exponential gravity model.
	}
	\label{fig:supp:predict_geo_bymodel}
\end{figure}

%
% Figure - Embedding predictions, by model
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance_emb_bymodel.pdf}
	\caption{
		\textbf{For embedding distance, the exponential gravity model is better.}
		Predictions of flux between organization pairs made using embedding distance the exponential gravity model (top), and the power-law decay model (bottom). 
		Boxplots show distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps $x = y$.
		Shown using all pairs of organization (\textbf{a-b}), domestic (\textbf{c-d}), and international only (\textbf{e-f}) mobility. 
		The exponential gravity model out-performs the power-decay model. 
	}
	\label{fig:supp:predict_emb_bymodel}
\end{figure}

%
% Figure - gravity correlation
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_correlations.pdf}
	\caption{
		\textbf{Embedding distance explains more variance for global, within, and between country flux than geographic distance.}
		\textbf{a.}
		Embedding distance (cosine distance between vectors) explains more of the total distance than does geographic distance (\textbf{b}).
		Red line is line of best fit.
		Black dots are mean flux across binned distances.
		Color indicates frequency of organization pairs within each hex bin.
		Results here are identical to those shown in Fig.~\ref{fig:gravity_performance}a and Fig.~\ref{fig:gravity_performance}.
		\textbf{c-d.}	embedding distance explains more variance when considering only pairs of organizations within the same country.
		\textbf{e-f.} embedding distance is more robust than geographic distance for pairs of organizations in different countries.
	}
	\label{fig:supp:gravity_correlation}
\end{figure}

@@

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_compare_fig.pdf}
	\caption{
		\textbf{SemRank hierarchy is robust.}
		\textbf{a.} Spearman correlation (n = 143) between Times prestige rank and SemAxis rank, by the number of high and bottom-ranked (geographically matched) universities aggregated to form the poles.
		Black points show spearman correlation using all organizations; white points show correlation using only universalizes not aggregated in the poles. 
		More universities improves performance, but quickly saturates after around five universities.
		\textbf{b - f.} Comparison between the Times and SemAxis ranks of universities, by number of universities used to define the poles (n).
		White points are those top and bottom 20 universities aggregated to define the ends of the axis.
		The grey box corresponds to the top 20 and bottom 20 ranks.
	  	PCC details the estimate from Spearman correlation between the two rankings.
	}
	\label{fig:supp:semaxis_compare}
\end{figure}

>>>>

Where the flux, $T_{ij}$, or number of authors co-affiliated with a pair organizations during the time period increases with the organization's sizes, $m_{i}$ and $m_{j}$, and decays as a function of their distance, $f(r_{ij})$.
We consider two distances, the geographic distance between the organizations as a baseline, and the \textit{embedding distance} defined as the cosine distance between the organization's vector in the trained embedding space.
We argue that the performance of a distance measure can be observed from their ability to explain expected mobility between organizations and to make predictions using the gravity model.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (ref SI.XYZ).

@@

The effective distance estimated by our embedding model explains and predicts scientific mobility best.
The amount of variance explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds for the variance of only domestic (Fig.~\ref{fig:supp:gravity_correlation}c) and international mobility flows (Fig.~\ref{fig:supp:gravity_correlation}d).
After fitting model parameters from the data, the embedding distance produces better predictions of expected flux than does geographic distance globally (Fig.~\ref{fig:gravity_performance}c--d) and for domestic (Fig.~\ref{fig:supp:gravity_correlation}e--f) and international mobility(Fig.~\ref{fig:supp:gravity_correlation}g--h).
Similarly, embedding distance outperforms ``diffusion''  distances such as those derived from calculating personalized-page rank scores on the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Geographic distance alone is a poor tool for understanding international scientific mobility---rather, by learning a distance directly from the data, it is possible to better explain, and better predict flux between organizations.

@@

%
% S1 TEXT
\newpage
\paragraph*{S1 Text}
\label{si:text:ppr_dist}
{\bf Network-based personalized page rank distances.}

@@

\newpage
%
% Figure - gravity correlation
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_correlations.pdf}
	\caption{
		\textbf{Embedding distance explains more variance for global, within, and between country flux than geographic distance.}
		\textbf{a.}
		Embedding distance (cosine distance between vectors) explains more of the total distance than does geographic distance (\textbf{b}).
		Red line is line of best fit.
		Black dots are mean flux across binned distances.
		Color indicates frequency of organization pairs within each hex bin.
		Results here are identical to those shown in Fig.~\ref{fig:gravity_performance}a and Fig.~\ref{fig:gravity_performance}.
		\textbf{c-d.}	embedding distance explains more variance when considering only pairs of organizations within the same country.
		\textbf{e-f.} embedding distance is more robust than geographic distance for pairs of organizations in different countries.
	}
	\label{fig:supp:gravity_correlation}
\end{figure}

@@

%
% Figure - SemAxis results
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/semaxis_compare_fig.pdf}
	\caption{
		\textbf{SemRank hierarchy is robust.}
		\textbf{a.} Spearman correlation (n = 143) between Times prestige rank and SemAxis rank, by the number of high and bottom-ranked (geographically matched) universities aggregated to form the poles.
		More universities improves performance, but quickly saturates after around five universities.
		\textbf{b - f.} Comparison between the Times and SemAxis ranks of universities, by number of universities used to define the poles (n).
		White points are those top and bottom 20 universities aggregated to define the ends of the axis.
		The grey box corresponds to the top 20 and bottom 20 ranks.
	  	PCC details the estimate from Spearman correlation between the two rankings.
	}
	\label{fig:supp:semaxis_compare}
\end{figure}
